Modelling the gas kinematics of an atypical Lyman alpha emitting compact dwarf galaxy

### 2017-03-13 paper/letter.tex
lya abbrev

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title[An atypical Lyman-$\alpha$ dwarf galaxy]{
Modelling the gas kinematics of an atypical Lyman-$\alpha$ emitting compact dwarf galaxy}
\author[J.E. Forero-Romero et al.]
{Jaime E. Forero-Romero $^{1}$ \thanks{je.forero@uniandes.edu.co},
Max Gronke$^2$, 
Maria Camila Remolina-Guti\'errez$^1$,
\newauthor
Nicol\'as Garavito-Camargo$^3$, 
Mark Dijkstra$^2$\\
%%
$^1$ Departamento de F\'isica, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia \\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
}

@@

\begin{abstract}
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ line emission that had evaded theoretical
interpretation so far. 
In this paper we explain these features by two different kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these kinematic conditions.
We find that both models independently require high velocities
(either a  a clump velocity dispersion of \sigmaclump\ with outflows of
\inftyclump\ or a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$)
consistent with a dynamical mass of at 
least a $10^{10}$ M$_{\odot}$, ten times larger than its estimated baryonic mass.   
We argue that a possible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
kinematics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the 
velocities found by our models, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
\end{abstract}

@@

Primordial galaxies have not been detected yet. 
However, dwarf star forming galaxies with a low metallicity content
are seen as templates to understand the early galaxy evolution process. 
Almost fifty years ago \citep{PartridgePeebles} it was realized that
young galaxies could be detected through a strong Lyman-$\alpha$ line
emission.

This theoretical prediction was only confirmed thirty years later on
distant, relatively young, not primordial, galaxies \citep{1998ApJ...498L..93D}.
Currently Lyman Alpha Emitting (LAE) galaxies are commonly targeted
in surveys. 
The presence of the Ly-$\alpha$ emission line provides confirmation of
the distance of a galaxy while provides clues about the stellar
population and inter-stellar medium conditions regulating the
Ly-$\alpha$ emission.
A careful clustering analysis of LAEs can also yield clues about their link
to dark matter halos \cite{2016ApJ...828....5M}.

The Ly-$\alpha$ emission line is not exclusive of distant galaxies. 
There are local Universe surveys that target Ly-$\alpha$ emission in nearby dwarf star forming galaxies  \citep{LARS}. 
The study of nearby LAE samples has allowed the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward \citep{2015ApJ...805...14R}. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
emerging Ly-$\alpha$ line
\cite{Harrington73,1991ApJ...370L..85N,LoebRybicki,2006ApJ...645..792T}. 
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of Ly-$\alpha$ observations
requires state-of-the-art Monte Carlo radiative transfer simulations.

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \citep{Thuan97} with two puzzling 
features: the line is symmetric and single peaked.
Usually the Ly-$\alpha$ line has an asymmetric single or double peak. 
These two special features in \tol\ cannot be explained quantitatively
tith conventional shell models
\citep{Mas-Hesse09,2015A&A...578A...7V,2015ApJ...812..123G}.

@@

The observed flux for the Lyman alpha line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \citep{Thuan97}
and a Equivalent Width of $70$\AA\ and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
\citep{Izotov04} which gives a Ly$\alpha$/H$\beta$ flux ratio of
4.9$\pm$0.1.
The Ly-$\alpha$ flux values correspond to luminosities of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$, which in turn
translates  into a star formation rate of $2.0$ M$_{\odot}$ yr$^{-1}$
using a standard conversion factor between luminosity and star
formation rate of $9.1\times 10^{-43}$ $L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.
% using http://tomdwelly.com/tools_fluxtolum.php
Comparing this ratio with the theoretical expectation from case B
recombination of $23.3$ \citep{Hummer1987} one can estimate an escape
fraction of $20$\% for Ly$\alpha$ radiation.
The bolometric UV luminosity is $9.43\pm1.94 \times 10^{8}$
L$_{\odot}$ as measured by GALEX, without any correction by extinction
and following the empirical relation by \cite{Kennicutt98}, this
corresponds to a star formation rate of $0.35\pm 0.05$ M$_{\odot}$
yr$^{-1}$. Its metallicity is $\sim Z_{\odot}/24$ as derived from
optical spectroscopy \citep{Izotov04}.

@@

\begin{table}
\begin{center}
\begin{tabular}{lc}\hline
$\alpha$(2000)$^{a}$ & 12h17min17.1s\\
$\delta$(2000)$^{b}$ & -28d02m32s\\
$l$, $b$ (deg) & 294, 34\\
$m_V$ & 17.5\\
  M$_V$ & -17.6\\ 
$v$(km s$^{-1}$) & 7795\\
Ly-$\alpha$ (erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$)& $8.1\times 10^{-14}$ \\
Ly-$\alpha$ Equivalent Width (\AA) & $70$\\
$21$cm (Jy km s$^{-1}$)& $<0.10$ \\\hline
\end{tabular}
\end{center}
\caption{Basic observational characteristics of TOL1214-277
  \citep{Thuan97}\label{obstable}} 
\end{table}

\begin{figure*}
\begin{center}
\includegraphics[width=0.55\textwidth]{CLARA-TOL-main.pdf}
\caption{Broad, single peaked and symmetric Ly-$\alpha$ emission of \tol.
  Dots correspond to the observational data. The line shows the results
of our best model from a full radiative transfer simulation both for
the rotation and multiphase models.
The observational points corresponds to the data as shown in
\citep{mashesse03}.}   
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.6\textwidth]{vinf_cl.pdf}
  \includegraphics[width=0.6\textwidth]{P_cl.pdf}
  \includegraphics[width=0.6\textwidth]{sigma_cl.pdf}
\caption{Results from the multiphase
  model.
  The left column corresponds to the parameter's integrated distributions for
  models with the lowest $\chi^2$ values (dotted line) compared against the
  parameter's integrated distributions (continous line) used as a prior.
  These are the only three parameters that show a significant statistical difference from
  the prior distributions, for all the other eleven parameters we
  cannot find any significant difference.
  The right column shows a scatter plot of $\chi^2/1000$
  and its corresponding physical parameters.
  Only the  25 models with the lowest $\chi^2$ are included.
  \label{multiphaseresults}
}  
\end{center}
\end{figure*}

@@

The rotation model corresponds to the work presented in
\citep{GaravitoCamargo2014} based on the Monte Carlo code
\texttt{CLARA} \citep{CLARA}. 
In that model the Ly-$\alpha$ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the HI line's
center optical  depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
Including the effect of dust only changes the overall line
normalization but not its shape.  
The results we report do not include any dust model.
Here we use an analytical solution that captures the most important
effects of rotation onto the \lya\ line as explained by
\citet{GaravitoCamargo2014} and allows us a full parameter space
exploration.

@@

A new observational test is needed to clarify the kinematical nature of
\tol. 
We suggest that integral field unit measurements 
resolving its spatial extent are up to the task. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer
\citep{2014Msngr.157...13B} with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
We note that archival Chandra X-ray archival
data \footnote{\url{http://cxc.harvard.edu/cda/}} does not show any
detection for \tol. This lack of observations also puts an upper
limit on the intensity of recent black hole activity.

>>>>

%=========================================================================
%		FRONT MATTER
%=========================================================================
\title[An atypical \lya\ dwarf galaxy]{
Modelling the gas kinematics of an atypical Lyman alpha emitting compact dwarf galaxy}
\author[J.E. Forero-Romero et al.]
{Jaime E. Forero-Romero $^{1}$ \thanks{je.forero@uniandes.edu.co},
Max Gronke$^2$, 
Maria Camila Remolina-Guti\'errez$^1$,
\newauthor
Nicol\'as Garavito-Camargo$^3$, 
Mark Dijkstra$^2$\\
%%
$^1$ Departamento de F\'isica, Universidad de los Andes, Cra. 1
  No. 18A-10 Edificio Ip, CP 111711, Bogot\'a, Colombia \\
$^2$ Institute of Theoretical Astrophysics, University of Oslo,
Postboks 1029 Blindern, NO-0315 Oslo, Norway.\\
$^3$ Department of Astronomy, University of Arizona, 933 North Cherry
Avenue, Tucson, AZ 85721, USA. 
}

@@

\begin{abstract}
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric \lya\ line emission that had evaded theoretical
interpretation so far. 
In this paper we explain these features by two different kinematic models: 
an interstellar medium composed by outflowing clumps with 
random motions and an homogeneous gaseous sphere undergoing solid body
rotation.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these kinematic conditions.
We find that both models independently require high velocities
(either a  a clump velocity dispersion of \sigmaclump\ with outflows of
\inftyclump\ or a bulk rotation of $348^{+75}_{-48}$ km s$^{-1}$)
consistent with a dynamical mass of at 
least a $10^{10}$ M$_{\odot}$, ten times larger than its estimated baryonic mass.   
We argue that a possible explanation for this excess of
dynamical mass is the presence of a supermassive black hole at the
center of \tol. 
This work demonstrates the importance of considering multiphase
kinematics and rotation among the possible conditions shaping the
\lya\ spectra of the first galaxies.  
Additionally, if future kinematic maps of \tol\ confirm the 
velocities found by our models, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
\end{abstract}

@@

Primordial galaxies have not been detected yet. 
However, dwarf star forming galaxies with a low metallicity content
are seen as templates to understand the early galaxy evolution process. 
Over fifty years ago \citep{PartridgePeebles} it was realized that
young galaxies could be detected through a strong \lya\ line
emission.

This theoretical prediction was only confirmed thirty years later on
distant, relatively young, not primordial, galaxies \citep{1998ApJ...498L..93D}.
Currently Lyman Alpha Emitting (LAE) galaxies are commonly targeted
in surveys. 
The presence of the \lya\ emission line provides confirmation of
the distance of a galaxy while provides clues about the stellar
population and inter-stellar medium conditions regulating the
\lya\ emission.
A careful clustering analysis of LAEs can also yield clues about their link
to dark matter halos \cite{2016ApJ...828....5M}.

The \lya\ emission line is not exclusive of distant galaxies. 
There are local Universe surveys that target \lya\ emission in nearby dwarf star forming galaxies  \citep{LARS}. 
The study of nearby LAE samples has allowed the study of other
indicators that might be more difficult to obtain for distant galaxies
such as morphology, dust attenuation, neutral hydrogen contents and
ionization state.

However, the physical interpretation of \lya\ observations is
not straightforward \citep{2015ApJ...805...14R}. 
This is due to the resonant nature of the \lya\ line. 
A \lya\ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
emerging \lya\ line
\cite{Harrington73,1991ApJ...370L..85N,LoebRybicki,2006ApJ...645..792T}. 
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of \lya\ observations
requires state-of-the-art Monte Carlo radiative transfer simulations.

\tol\ is a compact star forming dwarf galaxy that presents a
strong \lya\ emission \citep{Thuan97} with two puzzling 
features: the line is symmetric and single peaked.
Usually the \lya\ line has an asymmetric single or double peak. 
These two special features in \tol\ cannot be explained quantitatively
tith conventional shell models
\citep{Mas-Hesse09,2015A&A...578A...7V,2015ApJ...812..123G}.

@@

The observed flux for the \lya\ line is $\sim
8.1\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \citep{Thuan97}
and a Equivalent Width of $70$\AA\ and its H$\beta$ flux is 
$1.62\times 10^{-14}$ erg cm$^{-2}$ s$^{-1}$ \AA${-1}$
\citep{Izotov04} which gives a \lya/H$\beta$ flux ratio of
4.9$\pm$0.1.
The \lya\ flux values correspond to luminosities of
$L_{Ly\alpha}=2.2\times 10^{42}$ erg s$^{-1}$, which in turn
translates  into a star formation rate of $2.0$ M$_{\odot}$ yr$^{-1}$
using a standard conversion factor between luminosity and star
formation rate of $9.1\times 10^{-43}$ $L_{Ly\alpha}$ M$_{\odot}$ yr$^{-1}$. 
The absolute magnitude in the $V$ band translates into a luminosity of
$8.9\times 10^{8}$ L$_{\odot}$.
% using http://tomdwelly.com/tools_fluxtolum.php
Comparing this ratio with the theoretical expectation from case B
recombination of $23.3$ \citep{Hummer1987} one can estimate an escape
fraction of $20$\% for \lya\ radiation.
The bolometric UV luminosity is $9.43\pm1.94 \times 10^{8}$
L$_{\odot}$ as measured by GALEX, without any correction by extinction
and following the empirical relation by \cite{Kennicutt98}, this
corresponds to a star formation rate of $0.35\pm 0.05$ M$_{\odot}$
yr$^{-1}$. Its metallicity is $\sim Z_{\odot}/24$ as derived from
optical spectroscopy \citep{Izotov04}.

@@

\begin{table}
\begin{center}
\begin{tabular}{lc}\hline
$\alpha$(2000)$^{a}$ & 12h17min17.1s\\
$\delta$(2000)$^{b}$ & -28d02m32s\\
$l$, $b$ (deg) & 294, 34\\
$m_V$ & 17.5\\
  M$_V$ & -17.6\\ 
$v$(km s$^{-1}$) & 7795\\
\lya\ (erg cm$^{-2}$ s$^{-1}$ \AA$^{-1}$)& $8.1\times 10^{-14}$ \\
\lya\ Equivalent Width (\AA) & $70$\\
$21$cm (Jy km s$^{-1}$)& $<0.10$ \\\hline
\end{tabular}
\end{center}
\caption{Basic observational characteristics of TOL1214-277
  \citep{Thuan97}\label{obstable}} 
\end{table}

\begin{figure*}
\begin{center}
\includegraphics[width=0.55\textwidth]{CLARA-TOL-main.pdf}
\caption{Broad, single peaked and symmetric \lya\ emission of \tol.
  Dots correspond to the observational data. The line shows the results
of our best model from a full radiative transfer simulation both for
the rotation and multiphase models.
The observational points corresponds to the data as shown in
\citep{mashesse03}.}   
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
  \includegraphics[width=0.6\textwidth]{vinf_cl.pdf}
  \includegraphics[width=0.6\textwidth]{P_cl.pdf}
  \includegraphics[width=0.6\textwidth]{sigma_cl.pdf}
\caption{Results from the multiphase
  model.
  The left column corresponds to the parameter's integrated distributions for
  models with the lowest $\chi^2$ values (dotted line) compared against the
  parameter's integrated distributions (continous line) used as a prior.
  These are the only three parameters that show a significant statistical difference from
  the prior distributions, for all the other eleven parameters we
  cannot find any significant difference.
  The right column shows a scatter plot of $\chi^2$ (rescaled by a $1000$)
  and its corresponding physical parameters.
  Only the  25 models with the lowest $\chi^2$ are included.
  \label{multiphaseresults}
}  
\end{center}
\end{figure*}

@@

The rotation model corresponds to the work presented in
\citep{GaravitoCamargo2014} based on the Monte Carlo code
\texttt{CLARA} \citep{CLARA}. 
In that model the \lya\ photons are propagated 
within a spherical and homogeneous cloud of HI gas undergoing solid
body rotation.
The sphere is fully characterized by three parameters: the HI line's
center optical  depth $\tau$ measured from the center to its surface, the HI
temperature $T$, and the linear surface velocity $V_{\rm max}$.  
Including the effect of dust only changes the overall line
normalization but not its shape.  
The results we report do not include any dust model.
Here we use an analytical solution that captures the most important
effects of rotation onto the \lya\ line as explained by
\citet{GaravitoCamargo2014} and allows us a full parameter space
exploration.

@@

A new observational test is needed to clarify the kinematical nature of
\tol. 
We suggest that integral field unit measurements 
resolving its spatial extent are up to the task. 
\tol\ spans a region of $4$ arcseconds,
an instrument such as the Multi Unit Spectroscopic Explorer
\citep{2014Msngr.157...13B} with its
nominal $0.2$ arcseconds spatial sampling over a $1.0$ arcminute field
in wide-field mode could provide a coarse mapping of different
ionization lines to infer a kinematic map.
We note that archival Chandra X-ray archival
data \footnote{\url{http://cxc.harvard.edu/cda/}} does not show any
detection for \tol.
This lack of detection would also put an upper limit on the intensity
of recent black hole activity due to accretion.
